---
title: "Create a Sashimi Plot"
author: "James M. Ward"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
    toc: true
    toc_depth: 4
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Create a Sashimi Plot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
ragg_png = function(..., res = 192) {
  ragg::agg_png(..., res = res, units = "in")
}
knitr::opts_chunk$set(dev = "ragg_png", fig.ext = "png")
options("warn"=-1);
```

# Overview

This vignette is intended to describe how to create a Sashimi
plot using RNA-seq data.

## Outline

* `sashimiDataConstants()`: Prepare data inside an `environment`
which is used to prepare figures.
* `splicejamFigure()`: The core function to produce a Splicejam
sashimi plot figure.

Alternatively, open an R-shiny app:

* `launchSashimiApp()`


## Requirements

There are three basic requirements for a Sashimi plot:

1. **Gene-exon structure**: The easiest input is a GTF `.gtf` or GFF3 `.gff` file.
2. **RNA-seq coverage data**: Typically bigwig `.bw` files.
3. **Splice junction data**: Typically BED12 `.bed`, bigBed `.bb`, or
STAR `'SJ.out.tab'` formatted file.

Each type of data will be described below.


# Gene-exon structure

There are several sources of gene-exon structures:

## GTF File

GTF or GFF file, for example Gencode GTF (https://www.gencodegenes.com).

* A GTF file is sufficient to cover all gene-transcript-exon
requirements for Splicejam.
* When a GTF file and corresponding transcripts are used for
STAR alignment, and/or Salmon/Kallisto quantitation, we
recommend using the same file as input to Splicejam.
* The GTF should have the following fields defined:

   * 'gene_name': `character` gene symbol. Its values will be used
   to represent genes in the `data.frame` referred to as `'tx2geneDF'`.
   * 'transcript_id': `character` transcript identifier, for example
   `'ENST0000001.2'`. Its values will be used in `'tx2geneDF'` in
   column 'transcript_id'.
   * Note the 'gene_id' column is not typically used as the primary
   key for genes.

How the GTF file is used:

* The GTF or GFF file is used to prepare `'tx2geneDF'` which associates
transcripts `'transcript_id'` to genes `'gene_name'`, using
`makeTx2geneFromGtf()`.
It also parses other GTF entries in column 9 (annotations),
to include in the resulting `tx2geneDF` output `data.frame`.
* The GTF or GFF file is converted to `TxDb` annotation database
using `txdbmaker::makeTxDbFromGFF()`. This `TxDb` is saved in a
local file with extension `'.txdb'` for re-use if needed.
* The `TxDb` data is used to produce `exonsByTx`, and if the 'CDS'
field is defined in the GTF file it also produces `cdsByTx`. The
CDS is used to distinguish non-coding exons (thin exons) from coding
exons (wide exons) in the Splicejam sashimi plot.

## TxDb and tx2geneDF

A `TxDb` is typically produced using the Bioconductor package
'txdbmaker', see `txdbmaker::makeTxDb()` for various types
of input.

* Note that `TxDb` input is not sufficient alone, since Splicejam
requires 'gene_name' to be associated with 'transcript_id'.

   * Bioconductor provides several Txdb data packages
   (see BiocView
   https://bioconductor.org/packages/release/BiocViews.html#___TxDb).
   For example, to install the package for hg19 UCSC knownGenes TxDb,
   `BiocManager::install("TxDb.Hsapiens.UCSC.hg19.knownGene")`.
   * The TxDb specification includes 'gene_id' but not 'gene_name',
   instead it must be provided, or produced using Bioconductor
   organism annotation data, for example the Human annotation package
   is called `'org.Hs.eg.db'`.
   * Most Bioconductor packages define 'gene_id' using ENTREZID,
   however it is not clearly documented. For example
   `TxDb.Hsapiens.UCSC.hg38.knownGene` does not indicate that is uses
   ENTREZID, which is surprising because UCSC knownGene typically
   uses its own gene identifiers to enable them to provide gene
   loci which are not represented in the NCBI Gene release.
   Nonetheless, one can convert UCSC knownGene in hg38 using
   `org.Hs.eg.db::org.Hs.egSYMBOL` using something like this:
   `AnnotationDbi::mget("1", org.Hs.eg.db::org.Hs.egSYMBOL)`.

* Methods to produce `'tx2geneDF'` are described in `tximport::tximport()`,
which also includes references to other resources.
One straightforward approach is to use a GTF or GFF file
with `makeTx2geneFromGtf()`.

## GRangesList objects, and tx2geneDF

`GRangesList` objects 'exonsByTx', 'exonsByGene', and 'cdsByTx'
may be provided, for example by using `GenomicFeatures::exonsBy()`
and `GenomicFeatures::cdsBy()` or equivalent.

These objects are used to produce flattened exon models, for example
a flat exon model for transcript would combine `exonsByTx` with
`cdsByTx` to represent non-coding and coding regions as narrow
and wide exons, respectively. The output is `flatExonsByTx`.

Alternatively, `flatExonsByTx` can be provided directly, for
example it can be prepared using `flattenExonsBy()`.

The `flatExonsByTx` is used together with `tx2geneDF` to
prepare `flatExonsByGene`. Alternatively, it may be prepared
upfront for example using `flattenExonsBy()`.

> Note that
gene exons are numbered based upon the observed exon structure
of the transcripts provided. Therefore, when using `detectedTx`
to define a subset of detected transcripts, the exon numbers
may differ from the output if using all annotated transcripts.
This distinction was intentional, and motivated by numerous
Gencode comprehensive transcripts which may not be observed
in practice. However, one may prepare flattened exons using
all transcripts to override this decision.

* The `tx2geneDF` should be a `data.frame` with minimal columns
'transcript_id' and 'gene_name'.
These data associate transcripts to genes.
* `exonsByTx` should be named by 'transcript_id' as in `tx2geneDF`.
* `cdsByTx` should be named by 'transcript_id' as in `tx2geneDF`.
It is optional, and when provided it is used to produce narrow
and wide exons in the resulting gene-transcript-exon model.


## Use of detected transcripts

The recommended workflow is to use an appropriate
set of "detected transcripts" with Splicejam.
The motivation is to represent observed exon structure,
which is beneficial when used together with Gencode
comprehensive transcript data. The comprehensive
data often includes numerous extraneous transcripts
which may make it difficult to interpret isoform
changes in the given experiment.

Instead, using "detected transcripts" is helpful to
review scenarios where, for example, isoform A
switches to isoform B. There may be 12 other
potential isoforms without supporting data, and
they would not be helpful to assess the A-to-B
isoform switch.

Note that transcript quantitation is usually necessary
to define detected transcripts, for example using
Salmon or Kallisto.

Splicejam provides `splicejam::defineDetectedTx()` with
heuristics to help define detected transcripts, typically
using some combination of minimum counts,
minimum estimated TPM abundance, and
minimum percent max of each. It typically reduces the
total transcripts by about 60% the original size,
the vast majority of transcripts removed have zero
or near-zero quantitative values.

We found it beneficial to apply:

* Minimum count, to require some minimum plausible
supporting data.
* Minimum TPM abundance, to require some minimum
fractional transcript versus the overall observed
transcriptome.
* Percent max per gene, to require a minor
isoform to comprise at least 10% of the highest
group mean within the same sample group.
This option uses 'TPM' by default, so the
evaluation is done at 'molar'-like level and
is relatively independent of transcript length.
This filter helps remove extremely low abundance
transcripts from extremely high abundance genes,
for example GAPDH and ACTB may have 20 copies of
an obscure isoform, compared with over 120,000
counts of the predominant isoform.
The 20 copies may be biologically real, but in our
experience has not typically represented signal
useful to downstream analysis and visualization.


For additional information about transcript
quantitation, see:

Salmon (Patro et al 2017, https://combine-lab.github.io/salmon/)

Kallisto (Bray et al 2016, https://pachterlab.github.io/kallisto/about)


## TxDb Bioconductor package

This example uses Bioconductor data package
`TxDb.Mmusculus.UCSC.mm10.knownGene` which represents
UCSC known genes for the mouse `mm10` genome assembly.

First we assemble all exons by transcript, then
all CDS exons by transcript. We will use these
exons to assemble annotated gene models.

```{r, mm10_txdb, fig.height=5, fig.width=7}
options("warn"=-1);
suppressPackageStartupMessages(library(splicejam));
suppressPackageStartupMessages(library(jamba));
suppressPackageStartupMessages(library(kableExtra));
if (suppressPackageStartupMessages(require(TxDb.Mmusculus.UCSC.mm10.knownGene))) {

   # First obtain exons by transcript
   exonsByTxMm10 <- exonsBy(TxDb.Mmusculus.UCSC.mm10.knownGene,
      by="tx",
      use.names=TRUE);
   values(exonsByTxMm10@unlistData)$feature_type <- "exon";
   values(exonsByTxMm10@unlistData)$subclass <- "exon";
   
   # For added insight, obtain CDS exons by transcript (optional)
   cdsByTxMm10 <- cdsBy(TxDb.Mmusculus.UCSC.mm10.knownGene,
      by="tx",
      use.names=TRUE);
   values(cdsByTxMm10@unlistData)$feature_type <- "cds";
   values(cdsByTxMm10@unlistData)$subclass <- "cds";
}
```

The `TxDb` objects do not permit storing gene symbols,
so the example uses the Bioconductor annotation package
`org.Mm.eg.db` to assign `gene_name` values.
For values not found in `org.Mm.eg.db`, we use the
format `LOC#` to indicate the Entrez gene ID, for example
Entrez gene ID `1234` would be represented `LOC1234`.

```{r, tx2gene_mm10}
if (require(TxDb.Mmusculus.UCSC.mm10.knownGene)) {

   # Now prepare tx_name, gene_id, gene_name data.frame,
   # surprisingly difficult
   tx2geneMm10 <- suppressMessages(
      AnnotationDbi::select(TxDb.Mmusculus.UCSC.mm10.knownGene,
         keys(TxDb.Mmusculus.UCSC.mm10.knownGene, "GENEID"),
         columns=c("GENEID","TXNAME"),
         keytype="GENEID"
      )
   );
   tx2geneMm10 <- renameColumn(tx2geneMm10,
      from=c("GENEID", "TXNAME", "TXTYPE"),
      to=c("gene_id", "transcript_id", "transcript_type"));

   # add gene_name using org.Mm.eg.db
   if (suppressPackageStartupMessages(require(org.Mm.eg.db))) {
      gene_ids <- values(genes(TxDb.Mmusculus.UCSC.mm10.knownGene))$gene_id;
      gene_namesL <- mget(gene_ids,
         org.Mm.egSYMBOL,
         ifnotfound=NA);
      ## Convert list to vector taking the first gene_name each
      ## (All genes should only have one SYMBOL but there is no
      ## hard constraint so we should make absolutely sure to
      ## use only one value per gene.)
      gene_names <- unlist(heads(S4Vectors::List(gene_namesL), 1));
      ## Replace NA with LOC# format
      ## Note we use gsub() to ensure the data fits the expected format
      if (any(is.na(gene_names))) {
         gene_na <- which(is.na(gene_names));
         gene_names[gene_na] <- gsub("^([0-9]+)$", "LOC\\1",
            names(gene_names[gene_na]));
      }
      tx2geneMm10$gene_name <- gene_names[as.character(tx2geneMm10$gene_id)];
   } else {
      ## If we have no gene annotations, use the gene_id values
      tx2geneMm10$gene_name <- as.character(tx2geneMm10$gene_id);
   }
   # print the first 20 rows to show the content
   print(head(tx2geneMm10, 20));
}
```

Next we flatten transcript exons to the gene level using
the function `splicejam::flattenExonsBy()`.

We also flatten exons to the transcript level, which has
the effect of combining CDS exons alongside non-CDS exons.
The result is used when plotting transcript exon structures.

In both cases, the argument `genes=c("Gria1", "Ntrk2")` is
used as a convenience to produce data for only these two genes.

> We recommend using detected transcripts here, with the
argument `detectedTx`.

```{r, flatten_exons}
if (require(TxDb.Mmusculus.UCSC.mm10.knownGene)) {
   # flatten exons to the gene level
   # for speed, we will only process "Gria1", and "Ntrk2"
   flatExonsByGeneMm10 <- flattenExonsBy(exonsByTx=exonsByTxMm10,
      cdsByTx=cdsByTxMm10,
      by="gene",
      genes=c("Gria1", "Ntrk2"),
      tx2geneDF=tx2geneMm10,
      verbose=FALSE);

   # to be fancy, also flatten transcripts, to include CDS ranges   
   flatExonsByTxMm10 <- flattenExonsBy(exonsByTx=exonsByTxMm10,
      cdsByTx=cdsByTxMm10,
      tx2geneDF=tx2geneMm10,
      by="tx",
      genes=c("Gria1", "Ntrk2"));
}
```

#### Plot gene-exon models using gene2gg()

Once the gene-transcript-exon data is prepared, the exon
structure can be plotted using the function `splicejam::gene2gg()`:

```{r, gene_plot, fig.height=5, fig.width=7}
if (require(TxDb.Mmusculus.UCSC.mm10.knownGene)) {
   # Pull out Gria1
   grlGria1 <- flatExonsByGeneMm10[["Gria1"]];

   ## Plot a basic gene-exon structure
   ggGria1exons <- gene2gg(gene="Gria1",
      flatExonsByGene=flatExonsByGeneMm10,
      exonLabelSize=6);
   print(ggGria1exons + ggtitle("Gria1 exons"));

   ## Compare to the gene structure without compressing introns
   gg1full <- gene2gg(gene="Gria1",
      flatExonsByGene=flatExonsByGeneMm10,
      compressGaps=FALSE)
   print(gg1full);
   
   ## Plot a slightly more detailed gene-transcript-exon structure
   ggGria1exonsTx <- gene2gg(gene="Gria1",
      flatExonsByGene=flatExonsByGeneMm10,
      flatExonsByTx=flatExonsByTxMm10,
      tx2geneDF=tx2geneMm10);
   print(ggGria1exonsTx + ggtitle("Gria1 (compressed introns)"));
   
   ## Notice how difficult it is to see exon15 and exon16 are
   ## mutually exclusive exons
   gg2full <- gene2gg(gene="Gria1",
      flatExonsByGene=flatExonsByGeneMm10,
      flatExonsByTx=flatExonsByTxMm10,
      tx2geneDF=tx2geneMm10,
      compressGaps=FALSE);
   print(gg2full + ggtitle("Gria1 (uncompressed introns)"));

}
```


# Coverage and Junctions data

All coverage data, and junction data, should be described
in a `data.frame` `'filesDF'`. It should have the following
columns:

1. 'sample_id': Sample identifier, and in reality it should
represent sample groups even when multiple replicates are
available for each group. All files associated with each
'sample_id' are combined in strand-specific manner.
2. 'url': A web URL https link, or path to a local file,
or the `character` name of column `values()` when supplying
coverage as `GRanges` object `'covGR'`.
3. 'type': `character` string, one of:

   * 'bw': bigwig coverage file
   * 'junction': BED12, bigBed (using bed12), or
   `'SJ.out.tab'` file produced by STAR.
   * 'coverage_gr': coverage provided as `GRanges`
   as `'covGR'`. The names in `values()` must match
   the character string in 'url' for this row.

4. 'scale_factor': optional `numeric` scalar used to adjust
the magnitude of signal, applied by multiplying the scale
factor by the score.
The scale factor is intended to provide a method to normalize
data without creating a new coverage or junction file.

## Coverage data

The recommended input for coverage data is via bigWig
`'.bw'` or `'.bigwig'` files. BigWig data are imported
using the amazing package `cpp11bigwig::read_bigwig()`.

The bigwig files can be stored on the local file system,
or remote via https web URL. Local access is faster,
but remote access is fairly close due to 'cpp11bigwig',
and using memoise cache removes the distinction after
each gene is loaded at least once.

It is possible to provide coverage as `GRanges` data with
columns storing the coverage as `NumericList`, or other
object coercible to `list`.
The `GRanges` entries must exactly match the `names()`
of the corresponding `flatExonsByGene`, for example
`'Gria1_exon1'`, `'Gria1_exon2a'`, `'Gria1_exon2b'`, etc.
The colnames of `values()` should exactly match the
`'sample_id'` column values in the `filesDF` `data.frame`.

Note that coverage in either case should only represent one
strand, therefore samples are expected to have "+" and
"-" strand coverage data.

Use the 'farrisdata' package, Github `'jmw86069/farrisdata'`,
for an example of filesDF:


```{r farris-filesDF}
filesDF <- farrisdata::farris_sashimi_files_df;
jamba::kable_coloring(jamba::mixedSortDF(filesDF, byCols=c(1, 3)),
   row.names=FALSE,
   row_color_by="sample_id",
   colorSub=farrisdata::colorSub)
```

### Coverage in GRanges format

The `GRanges` object below uses an example set of exons,
`test_exon_wide_gr`, and corresponding coverage `GRanges`
is `test_exon_wide_gr`. Note that both share the same
names and genomic coordinates.

There is no coverage defined for introns.

```{r, test_cov_data}
data(test_exon_wide_gr);
test_exon_wide_gr;
data(test_cov_wide_gr);
test_cov_wide_gr;
```

## Junctions data

Splice junction data are provided as a link to
local file, or remote web https URL.
Entries can be in BED6 format, BED12 format, or
'SJ.out.tab' format produced by STAR.

For BED formats, data is accepted as BED or bigBed,
although the bigBed format is substantially faster.

Splice junction data can be provided in two forms:

* Files in `BED6` or `BED12` format, accessible either as a file or web URL.
The `BED6` format uses the score column to count reads, and uses
the genomic coordinates to indicate the splice junction span.
* The `BED12` format is inverted, so there is one base
on each side, and the junction is in the middle.
In this case, the one base is trimmed from each side,
and the internal region is used to indicate the junction.
This approach works best when the overhang is fixed at 1 base
on each side.

For more insight into creating a Sashimi plot manually,
see the help for data `splicejam::test_junc_wide_gr`.


# Splicejam Data Environment

With the necessary data available, you create
a Splicejam `environment` to contain the data.

`sjenv <- sashimiDataConstants(...)`

By default, if you provide no supporting data, the
default argument `empty_uses_farrisdata=TRUE` means
it will prepare supporting data for the
Farris et al 2019 publication.

However, a small subset of data are available with the
test data `sjenvtest`.

The two most common types of input are:

1. GTF file.
2. TxDb file, and tx2geneDF `data.frame`.

For example, the typical pattern:

```{r create-sjenv, eval=FALSE}
sjenv <- new.env();
sjenv <- sashimiDataConstants(envir=sjenv,
   empty_uses_farrisdata=FALSE,
   filesDF=filesDF,
   gtf=gtf);
```

You can use `jamba::sdim()` to describe the data.
In this case, we will use `sjenvtest` as the `environment`:

```{r sdim-sjenvtest}
data(sjenvtest)
jamba::sdim(sjenvtest)
```

## Splicejam Figure

The core function `splicejamFigure()` prepares the
multi-panel plot for a given gene.

To see progress, enable it with `progressr::handlers(global=TRUE)`.

```{r splicejamFigure-1, fig.height=10, fig.width=14, alt="Sashimi plot showing the gene 'Gria1'."}
# optionally enable progress bar
# progressr::handlers(global=TRUE)

data(sjenvtest)
sjfig <- splicejamFigure(sjenv=sjenvtest,
   gene="Gria1")
```

It **does not enable memoise** by default, however
it is strongly recommended.

```{r splicejamFigure-2, eval=FALSE}
data(sjenvtest)
# Enable memoise
sjfig <- splicejamFigure(sjenv=sjenvtest,
   use_memoise=TRUE,
   gene="Gria1")
```

Note that `splicejamFigure()` does the work of calling
`prepareSashimi()`, `plotSashimi()`, `gene2gg()`,
then assembles the panels using `patchwork` by default.
Various customizations can be passed via '...' especially
with `plotSashimi()` arguments.


### Zoom into specific exons

The `splicejamFigure()` arguments include two methods
to zoom to specific exons.

1. 'exon_range': `character` vector with two exons, for
example: `use_exon_range=c('Gria1_exon13', 'Gria1_exon16')`
2. 'display_coords': `integer` vector with start and end
positions. Note the coordinates must be valid for the
gene being displayed.

```{r splicejamFigure-3, fig.height=10, fig.width=14, alt="Sashimi plot showing the gene 'Gria1', now zoomed in to show exon 13 through exon 16."}
data(sjenvtest)
sjfig_zoom <- splicejamFigure(sjenv=sjenvtest,
   use_exon_range=c('Gria1_exon13', 'Gria1_exon16'),
   use_memoise=TRUE,
   gene="Gria1")
```

### Adjust layout

You can adjust the layout, for example to impose
two columns use `layout_ncol=2`

```{r splicejamFigure-4, fig.height=12, fig.width=12, alt="Sashimi plot showing the gene 'Gria1', showing exon 13 through exon 16, and now with two-column layout."}
data(sjenvtest)
sjfig_zoom_2col <- splicejamFigure(sjenv=sjenvtest,
   use_exon_range=c('Gria1_exon13', 'Gria1_exon16'),
   layout_ncol=2,
   use_memoise=TRUE,
   gene="Gria1")
```


## Splicejam Shiny App

When the `sjenv` data `environment` is prepared,
you can run the Splicejam shiny app:

```{r splicejam-app, eval=FALSE}
data(sjenvtest)
launchSashimiApp(envir=sjenvtest)
```


### Start the Shiny app on a different port

Two common options to customize are 'host'
and 'port'. By using host='0.0.0.0' it makes
the Sashimi app accessible to other computers,
provided they can see your computer IP address.

Note that in general, the only computers that
are able to see your computer IP address are
those on the same network as your computer.
That goes for company or school intranet, or
within your home network.

A custom 'port' makes it easier to use a consistent
port number, otherwise it is more or less random.
On most computers, it requires some administrative
permission to use a port lower than 8000, so
a common recommendation is `port='8080'` or
something easy to remember.

```{r shiny_port, eval=FALSE}
launchSashimiApp(
   empty_uses_farrisdata=FALSE,
   envir=sjenv,
   options=list(
      port=8080,
      host="0.0.0.0"))
```

### Custom options

Several elements in the Splicejam app can
be set as defaults.

* **default_gene**: `character` gene symbol for the initial figure.
* **layout_ncol**: default number of columns for the panel layout.
* **panel_height**: height for each panel in pixels, default 200.
* **gene_panel_height**: height for the gene panel, default is 400.
* **min_junction_reads**: `numeric` minimum junction reads.
* **share_y_axis**: `logical` whether to use shared (fixed) y-axis
range across all panels, default FALSE.
* **use_exon_range**: `character` vector with first and last exon
to include by default.
* **gene_coords_default**: initial `integer` coordinate range,
default NULL shows the whole gene.
* **label_junctions**: `logical` whether to display junction counts,
default TRUE.
* **show_gene_model**: `logical` whether to show the gene-exon model,
default TRUE.
* **font_sizing**: `character` label matching one of the options shown,
default is "Default". Other options range from "-4 smaller" to
"+4 larger".
* **exon_font_sizing**: `character` label matching one of the options
shown, default is "Default". See comment for **font_sizing**.
* **junction_alpha**: `numeric` default alpha transparency, default
0.7 is 70% opaque.
* **junction_arc_factor**: `character` string matching one of the
options shown, default is "Default". Options range from
"-2 flat" to "+3 higher".
* **junction_arc_minimum**: `numeric` minimum arc height per junction,
default 500 ensures the arc is at least 500 y-axis units high.
* **aboutExtra**: `character` or `htmlwidgets` formatted tags
suitable to display as HTML, shown in the "About" tab to describe
the data being displayed. It is normally defined inside
`sashimiAppConstants()`.

For example:

```{r Splicejam-app-2, eval=FALSE}
data(sjenvtest)
# or prepare your own environment
# sjenv <- sashimiDataConstants()

sjenvtest$default_gene <- "Ntrk3"
sjenvtest$panel_height <- 150;
sjenvtest$gene_panel_height <- 300;
sjenvtest$use_exon_range <- c("Ntrk3_exon1", "Ntrk3_exon23a");

launchSashimiApp(envir=sjenvtest)

```
