---
title: "Create a Sashimi Plot"
author: "James M. Ward"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
    toc: true
    toc_depth: 4
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Create a Sashimi Plot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
ragg_png = function(..., res = 192) {
  ragg::agg_png(..., res = res, units = "in")
}
knitr::opts_chunk$set(dev = "ragg_png", fig.ext = "png")
options("warn"=-1);
```

# Overview

This vignette is intended to describe how to create a Sashimi
plot using RNA-seq data.

1. Define coverage and junction data: `filesDF`
2. Provide gene-exon data: `sjenv`
3. Create Splicejam figures.

# Data Requirements

The process:

1. **Splicejam environment**:

   * **TxDb** from Bioconductor, or your own.
   * **GTF file** with specific annotations.

2. **Create `filesDF`**:

   * **RNA-seq coverage data**: bigwig `.bw` files.
   * **Splice junction data**: `.bed`, bigBed `.bb`, or STAR `'SJ.out.tab'`.


# Create the Splicejam data environment

Create the Splicejam `environment`
using one of the two paths described above.

For testing, we provide `sjenvtest` which includes data
for mouse genes: `'Gria1'` and `'Ntrk3'`.

See [Create `filesDF`](#Create `filesDF`) to provide `filesDF` at this step.

## GTF input

```{r create-sjenv, eval=FALSE}
sjenv <- new.env();
sjenv <- sashimiDataConstants(envir=sjenv,
   empty_uses_farrisdata=FALSE,
   filesDF=filesDF,
   gtf=gtf);
```

> Note:
If you call `sashimiDataConstants()` with no
additional arguments, the default
`empty_uses_farrisdata=TRUE` will try to use
data from Farris et al 2019 via the Github
R package `'jmw86069/farrisdata'`.

## TxDb input

```{r create-sjenv-txdb, eval=FALSE}
sjenv <- new.env();
sjenv <- splicejamDataFromTxdb(
   txdb=TxDb.Mmusculus.UCSC.mm10.knownGene::TxDb.Mmusculus.UCSC.mm10.knownGene,
   ann_lib="org.Mm.eg.db",
   filesDF=filesDF);
```

* The `TxDb` package must already be installed, in this case
`'TxDb.Mmusculus.UCSC.mm10.knownGene'`.
* If `ann_lib` is not provided, it will be determined
by the 'Organism' found in the `TxDb` metadata.


## Review sjenv

You can use `jamba::sdim()` to describe the data.
In this case, use `sjenvtest`:

```{r sdim-sjenvtest-show, eval=FALSE}
jamba::sdim(sjenvtest)
```

```{r sdim-sjenvtest, echo=FALSE}
library(splicejam)
data(sjenvtest)
jamba::kable_coloring(
   jamba::sdim(sjenvtest),
   caption="Summary of objects in 'sjenv', with length or dimensions as relevant to each object.")
```


# Create the Splicejam figure

## splicejamFigure

The core function `splicejamFigure()` prepares the
multi-panel plot for a given gene.

Optionally, enable a progress bar with
`progressr::handlers(global=TRUE)`.

```{r splicejamFigure-1, fig.height=10, fig.width=14, alt="Sashimi plot showing the gene 'Gria1'."}
# optionally enable progress bar
# progressr::handlers(global=TRUE)

data(sjenvtest)
sjfig <- splicejamFigure(sjenv=sjenvtest,
   gene="Gria1")
```

**We recommend using memoise to cache data**,
with argument `use_memoise=TRUE` shown below.

```{r splicejamFigure-2, eval=FALSE}
data(sjenvtest)
# Enable memoise
sjfig <- splicejamFigure(sjenv=sjenvtest,
   use_memoise=TRUE,
   gene="Gria1")
```

`splicejamFigure()` does the work of calling:

* `prepareSashimi()` - coverage and junction data
* `plotSashimi()` - ggplot2 object
* `gene2gg()` - gene-transcript-exon model
* `patchwork` - multipanel figure

Various customizations can be passed to these functions
using '...'. See `plotSashimi()`.


## Zoom by exon names

The `splicejamFigure()` arguments include two methods
to zoom to specific exons.

1. `'exon_range'`: `character` vector with two exons, for
example: `use_exon_range=c('Gria1_exon13', 'Gria1_exon16')`
2. `'display_coords'`: `integer` vector with start and end
positions. Note the coordinates must be valid for the
gene being displayed.

```{r splicejamFigure-3, fig.height=10, fig.width=14, alt="Sashimi plot showing the gene 'Gria1', now zoomed in to show exon 13 through exon 16."}
data(sjenvtest)
sjfig_zoom <- splicejamFigure(sjenv=sjenvtest,
   use_exon_range=c('Gria1_exon13', 'Gria1_exon16'),
   use_memoise=TRUE,
   gene="Gria1")
```


## Adjust column layout

You can adjust the layout, for example to impose
two columns use `layout_ncol=2`.

Supply `sample_id` in a specific order to control
the order each panel is displayed.


```{r splicejamFigure-4, fig.height=12, fig.width=12, alt="Sashimi plot showing the gene 'Gria1', showing exon 13 through exon 16, and now with two-column layout."}
data(sjenvtest)
sjfig_zoom_2col <- splicejamFigure(sjenv=sjenvtest,
   use_exon_range=c('Gria1_exon13', 'Gria1_exon16'),
   layout_ncol=2,
   use_memoise=TRUE,
   gene="Gria1")
```


# Create filesDF

Create `'filesDF'` to provide coverage and junctions data.

An example of the `data.frame` format is shown below:

```{r example_filesDF, echo=FALSE, results='asis'}
#| tbl-cap: A data.frame is shown as an example of the format required by 'filesDF'.
example_filesDF <- data.frame(check.names=FALSE,
   "sample_id"=rep(c('Sample A', 'Sample B', '...'), c(3, 3, 1)),
   "url"=c('./Sample_A_pos.bw', './Sample_A_neg.bw', './Sample_A_SJ.out.tab',
      './Sample_B_pos.bw', './Sample_B_neg.bw', './Sample_B_SJ.out.tab', '...'),
   "type"=c('bw', 'bw', 'junction', 'bw', 'bw', 'junction', '...'),
   "scale_factor"=rep(c("1", "..."), c(6, 1))
)
use_colors <- c(bw='gold', junction='orange',
   "Sample A"='slateblue',
   "Sample B"='aquamarine1',
   jamba::nameVector('indianred', example_filesDF$url[1]),
   jamba::nameVector('indianred', example_filesDF$url[4]),
   jamba::nameVector('lavender', example_filesDF$url[3]),
   jamba::nameVector('lavender', example_filesDF$url[6]),
   jamba::nameVector('skyblue', example_filesDF$url[2]),
   jamba::nameVector('skyblue', example_filesDF$url[5]))
jamba::kable_coloring(example_filesDF,
   caption="A data.frame is shown as an example of the format required by 'filesDF'.",
   colorSub=use_colors)
```


## Coverage files

Provide files or web URLs to BigWig `.bw` or `.bigWig` files.

* BigWig data are imported using the amazing
`cpp11bigwig::read_bigwig()`, which works great
on Linux, MacOS, and Windows.
* Coverage can only represent one strand at a time.
Files with `'neg'` or `'-'` or `'minus'` in the
filename are assumed to be negative strand.

There is a provision to use `GRanges` data,
beyond the scope of this vignette.


## Junction files

Provide files or web URLs to files in one of these formats:

1. BED or BED12 file, in text format. BED12 format should
use two 1-base "overhang" on each side, with junction
defined by the gap.
2. SJ.out.tab format as produced by STAR alignment.
3. Bigbed `.bb` or `.bigBed` indexed file, best for
remote files.

Options 1 and 2 may use any compressed format such as `.gz`
which is readable by `data.table::fread()`.

The 'name' column is used when names are entirely numeric,
otherwise the 'score' column is used. The reason is that
'score' is limited to `1000` in bigBed `.bb` files.


## Scale Factor

The optional column `'scale_factor'` permits
`numeric` values used as scalar, to be multiplied by the
score for the corresponding file.

The scale factor is intended to help normalize signal
without needing to modify the coverage or junction file.

By default, all values are `1`.


## Example filesDF

Use the 'farrisdata' package, Github `'jmw86069/farrisdata'`,
for an example of filesDF:

```{r farris-filesDF-show, eval=FALSE}
farrisdata::farris_sashimi_files_df;
```

```{r farris-filesDF, echo=FALSE, results='asis'}
filesDF <- jamba::mixedSortDF(
   farrisdata::farris_sashimi_files_df,
   byCols=c(1, 3))[, 1:4]
filesDF$url <- file.path(".", basename(filesDF$url))
jamba::kable_coloring(jamba::mixedSortDF(filesDF, byCols=c(1, 3)),
   row.names=FALSE,
   row_color_by="sample_id",
   colorSub=farrisdata::colorSub)
```



# More about gene-exon structure

There are two main sources of gene-exon structures.

Some guidance on which path to follow:

1. **GTF Path**:
If you used a GTF file for data processing and
need to use the 'gene_name' values exactly as defined
in that GTF file, choose the GTF path below.
2. **TxDb Path**:
In all other cases, choose a suitable `TxDb` from
Bioconductor, or provide your own custom `TxDb` file.

**TxDb is the "easiest" path to use Splicejam.**

In both cases, the input provides a `TxDb` transcript
database.

* **GTF path** defines 'gene_name' from the GTF file.
* **TxDb path** uses Bioconductor 'SYMBOL', or you can
provide `tx2geneDF` yourself

> In future, `EnsDb` data format will also be supported.
File an issue on Github to request this change.


## GTF File

The GTF file should provide column 9 annotations:
The standard GTF input is based on the
Gencode GTF format (https://www.gencodegenes.com).

* 'gene_name': `character` gene symbol, used as the primary gene
identifier in Splicejam.
* 'gene_id': `character` identifier, used in `TxDb` data but
not used in Splicejam otherwise.
* 'transcript_id': `character` transcript identifier used
to associate transcript to gene.

It should provide features: 'gene', 'transcript', 'exon'.
'CDS' is used if available.


## TxDb

A `TxDb` transcript database may be obtained from
Bioconductor, such as `'TxDb.Hsapiens.UCSC.hg38.knownGene'`,
or can be provided as a file.

For more information about creating a TxDb, see `txdbmaker::makeTxDb()`
which supports using
a GTF or GFF3 file; BioMart; UCSC; EnsEMBL; or `GRanges` data.

The `TxDb` is used to create `'tx2geneDF'` as a `data.frame`,
using Bioconductor annotation data,
which associates `'transcript_id'` to `'gene_name'`.
You may also supply a custom `'tx2geneDF'`.

The default uses `TxDb` 'Organism' to determine a
suitable annotation library, or use argument `'ann_lib'`.
Common examples:

* `'org.Hs.eg.db'` for *Homo sapiens* (Human)
* `'org.Mm.eg.db'` for *Mus musculus* (Mouse)

For other methods to produce `'tx2geneDF'`, see
`tximport::tximport()`.


## Detected genes or detected transcripts

You can supply `'detectedTx'` with specific
`'transcript_id'` entries, which will subset the
transcripts used to define flattened gene-exon models.
It is possible to apply `'detectedTx'` either before
or after the initial processing step.

When providing `'detectedTx'` upfront, the flattened
gene-exon model will only contain exons from those
transcripts. The exons are numbered at this stage,
and may differ depending which transcripts are used.

If you use all transcripts upfront, then `'detectedTx'`
later, the gene-exon model will be numbered using
all transcripts, while the figure will only display
transcripts defined in `'detectedTx'`.


### Heuristics for detected transcripts

See `defineDetectedTx()` to help define
detected transcripts, intended to work with
data produced by `tximport::tximport()`.
It often reduces the total transcripts to 60% the original size.

The heuristics:

* **Minimum count**: require a minimum plausible
supporting data.
For example, at least 8 read counts.
* **Minimum TPM abundance**: require minimum
transcript abundance in TPM units
For example, at least 0.1 TPM.
* **Percent max per gene**: using TPM or read count.
TPM is recommended as a per-molecule metric.
For example, the least abundant isoform must be
at least 10% of the most abundant isoform.
This step helps remove extremely low abundance
isoforms which often appears as noise in transcript
quantitation results.
See genes like "GAPDH" and "ACTB".

For additional information about transcript
quantitation, see:

Salmon (Patro et al 2017, https://combine-lab.github.io/salmon/)

Kallisto (Bray et al 2016, https://pachterlab.github.io/kallisto/about)


## Bonus: Plot gene-exon models

Once the Splicejam `environment` `sjenv` is prepared,
you can plot the gene model and review the exon numbering
using `splicejam::gene2gg()`.

```{r gene-plot, fig.height=4, fig.width=10, alt="Gria1 gene-transcript-exon model showing introns with fixed, compressed width compared to exon widths."}
data(sjenvtest)
ggGria1 <- gene2gg(
   sjenv=sjenvtest,
   gene="Gria1");
print(ggGria1 + ggplot2::ggtitle("Gria1 (compressed introns)"));
```

The default above will compress the introns to a fixed
width, to help see each exon in detail.

You can show introns at full size using
`compressGaps=FALSE`:

```{r gene-plot-wide, fig.height=4, fig.width=10, alt="Gria1 gene-transcript-exon model showing introns at the full width."}
ggGria1_wide <- gene2gg(
   sjenv=sjenvtest,
   compressGaps=FALSE,
   gene="Gria1");
print(ggGria1_wide + ggplot2::ggtitle("Gria1 (full introns)"));

```




# Splicejam Shiny app

Using the `sjenv` data `environment`,
you can run the Splicejam shiny app with
`launchSashimiApp(envir=sjenv)`.
It will open within RStudio or Positron,
or will open the configured web browser
from an R console.

```{r splicejam-app, eval=FALSE}
data(sjenvtest)
launchSashimiApp(envir=sjenvtest)
```


## Change the host and port

* Use `host='0.0.0.0'` to make
the Sashimi app accessible to other computers
on your network.
* Use `host='localhost'` or `host='127.0.0.1'`
to limit access to your computer.
* If `'port'` is not defined, it chooses at random.
* If `'port'` is defined, choose 8000 or higher.


To make the app available outside your network,
consider https://connect.posit.cloud/.
Do not make your computer accessible to the internet.

An example:

```{r shiny_port, eval=FALSE}
launchSashimiApp(
   empty_uses_farrisdata=FALSE,
   envir=sjenv,
   options=list(
      port=8080,
      host="0.0.0.0"))
```

## Other custom options

The custom options must be added to the
`sjenv` in the `environment`. For example,
you can set two-column layout by default with:

`sjenv$layout_ncol=2`

These options can be customized by the user,
and only change the initial default condition.

Other options:

```{r other-shiny-options, results='asis', echo=FALSE}
params_df <- data.frame(
  Parameter = c(
    "default_gene", "layout_ncol", "panel_height", "gene_panel_height",
    "min_junction_reads", "share_y_axis", "use_exon_range", "gene_coords_default",
    "label_junctions", "show_gene_model", "font_sizing", "exon_font_sizing",
    "junction_alpha", "junction_arc_factor", "junction_arc_minimum", "aboutExtra"
  ),
  Description = c(
    "character: gene symbol for the initial figure.",
    "integer: number of columns for the panel layout.",
    "integer: pixel height for each panel, default 200.",
    "integer: pixel height for the gene panel, default 400.<br>\nUsually about 2x the panel_height.",
    "numeric: minimum junction reads, all others are hidden.<br>\nDefault 10.",
    "logical: whether to share the y-axis across all panels.<br>\nDefault FALSE.",
    "character: vector with first and last exons, default NULL.",
    "integer: coordinate range, default NULL.",
    "logical: whether to display junction counts, default TRUE.",
    "logical: whether to show the gene-exon model, default TRUE.<br>\nNot yet re-implemented.",
    "character: label matching one of the options shown.<br>\nDefault 'Default'. Range from '-4 smaller' to '+4 larger'.",
    "character: label matching one of the options shown.<br>\nDefault 'Default'. See font_sizing.",
    "numeric: default alpha transparency, default 0.7.<br>\n(70% opaque)",
    "character: string matching one of the options shown.<br>\nDefault 'Default'. Range from '-2 flat' to '+3 higher'.",
    "numeric: minimum arc height per junction, default 500.",
    "character or htmlwidgets: text suitable to display as HTML.<br>\nShown in the 'About' tab to describe experimental data."
  )
)
suppressPackageStartupMessages(library(magrittr))
jamba::kable_coloring(
   params_df) %>%
   kableExtra::column_spec(column=1, bold=TRUE)

```

For example:

```{r Splicejam-app-2, eval=FALSE}
data(sjenvtest)
# or prepare your own environment
# sjenv <- sashimiDataConstants()

sjenvtest$default_gene <- "Ntrk3"
sjenvtest$panel_height <- 150;
sjenvtest$gene_panel_height <- 300;
sjenvtest$use_exon_range <- c("Ntrk3_exon1", "Ntrk3_exon23a");

launchSashimiApp(envir=sjenvtest)

```
