---
title: "Create a Sashimi Plot"
author: "James M. Ward"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
    toc: true
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Create a Sashimi Plot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options("warn"=-1);
```

## Overview

This vignette is intended to describe how to create a Sashimi
plot using RNA-seq data.

### Requirements

There are three basic requirements for a Sashimi plot:

1. Gene-exon structure
2. RNA-seq coverage data
3. exon-exon splice junction data


## Gene-exon structure

There are several sources of gene-exon structures:

* GTF or GFF file, for example Gencode GTF (https://www.gencodegenes.com).

    * If a GTF file is used in upstream processing, it is recommended
    to use that same GTF or GFF file to create a `GenomicFeatures::TxDb`
    object with `GenomicFeatures::makeTxDbFromGFF()`.
    * The function `splicejam::makeTx2geneFromGtf()` creates a
    `data.frame` with cross-reference data for `"transcript_id"` and
    `"gene_name"`. This step makes GTF a more reliable workflow
    than the `TxDb` workflow, since the `TxDb` workflow
    relies upon Bioconductor annotations to obtain `"gene_name"`,
    and is sometimes lacking.

* Txdb transcript database, provided by `GenomicFeatures::TxDb`

    * Bioconductor provides several Txdb data packages
    (see BiocView
    https://bioconductor.org/packages/release/BiocViews.html#___TxDb).
    For example, to install the package for hg19 UCSC knownGenes TxDb,
    `BiocManager::install("TxDb.Hsapiens.UCSC.hg19.knownGene")`.
    * The `TxDb` specification does not include `"gene_name"` which
    therefore needs to be added using Bioconductor annotation packages,
    such as `org.Hs.eg.db` (Human, Homo sapiens) or `org.Mm.eg.db`
    (Mouse, Mus musculus). This step is almost always a little lossy,
    seen as `TxDb` entries whose `"transcript_id"` do not
    have corresponding`"gene_id"` nor `"gene_name"` entries.
    Ability to fetch annotation directly from UCSC is currently not
    straightforward.

* GRanges or GRangesList objects, already assembled by other methods.

    * For example `rtracklayer::import()` has several methods that can
    import BED files, or GTF files. Typically, BED format Ultimately the requirement is a
    disjoint (non-overlapping) set of exon GRanges, pulled into a
    GRangesList whose names are `"gene_name"`. This `"gene_name"`
    is mainly used as a figure caption, though it can be integrated
    with transcript exons.

The requirement for Sashimi plots is to flatten exons
to disjoint (non-overlapping) exons across the gene
region of interest. The exons are used to determine
the genomic regions to use for sequence coverage, and
splice junction reads.

The exons are also used to define the genome coordinate
transformation function used to compress introns to more
manageable (smaller) size.


### Use of detected transcripts

The recommended workflow is to use detected transcripts
derived from `splicejam::defineDetectedTx()` or some
other means. This step typically reduces the total
transcripts by 60%, which greatly simplifies the
resulting gene structure, and focuses the gene structure
on the experimental data.

Gene transcriptome data resources are increasingly comprehensive,
as they are derived from numerous tissue samples and cell types.
In ideal cases, the data should represent the superset
of possible gene transcripts, of which only a fraction are
experimentally observed.

In addition, we observed
questionable transcript annotations, with no apparent
supporting data -- possible artifacts of an automated
annotation pipeline. While these apparent errors are
found in low percentage, their presence can completely
disrupt a gene structure, sometimes combining several
genes (70+) into one extremely long gene locus longer than
5 megabases.

As a practical matter, we found it beneficial that
many of these transcripts get removed during the
`splicejam::defineDetectedTx()` process, conveniently
simplifying the resulting gene-exon models. Note this
process currently uses data from an analysis tool such as
Kallisto (Bray et al 2016, https://pachterlab.github.io/kallisto/about)
or
Salmon (Patro et al 2017, https://combine-lab.github.io/salmon/)
since these tools estimate transcript isoform abundance.

### Starting with TxDb

This example uses Bioconductor data package
`TxDb.Mmusculus.UCSC.mm10.knownGene` which represents
UCSC known genes for the mouse `mm10` genome assembly.

First we assemble all exons by transcript, then
all CDS exons by transcript. We will use these
exons to assemble annotated gene models.

```{r, mm10_txdb, fig.height=5, fig.width=7}
options("warn"=-1);
suppressPackageStartupMessages(library(splicejam));
suppressPackageStartupMessages(library(jamba));
suppressPackageStartupMessages(library(kableExtra));
if (suppressPackageStartupMessages(require(TxDb.Mmusculus.UCSC.mm10.knownGene))) {

   # First obtain exons by transcript
   exonsByTxMm10 <- exonsBy(TxDb.Mmusculus.UCSC.mm10.knownGene,
      by="tx",
      use.names=TRUE);
   values(exonsByTxMm10@unlistData)$feature_type <- "exon";
   values(exonsByTxMm10@unlistData)$subclass <- "exon";
   
   # For added insight, obtain CDS exons by transcript (optional)
   cdsByTxMm10 <- cdsBy(TxDb.Mmusculus.UCSC.mm10.knownGene,
      by="tx",
      use.names=TRUE);
   values(cdsByTxMm10@unlistData)$feature_type <- "cds";
   values(cdsByTxMm10@unlistData)$subclass <- "cds";
}
```

The `TxDb` objects do not permit storing gene symbols,
so the example uses the Bioconductor annotation package
`org.Mm.eg.db` to assign `gene_name` values.
For values not found in `org.Mm.eg.db`, we use the
format `LOC#` to indicate the Entrez gene ID, for example
Entrez gene ID `1234` would be represented `LOC1234`.

```{r, tx2gene_mm10}
if (require(TxDb.Mmusculus.UCSC.mm10.knownGene)) {

   # Now prepare tx_name, gene_id, gene_name data.frame,
   # surprisingly difficult
   tx2geneMm10 <- suppressMessages(
      AnnotationDbi::select(TxDb.Mmusculus.UCSC.mm10.knownGene,
         keys(TxDb.Mmusculus.UCSC.mm10.knownGene, "GENEID"),
         columns=c("GENEID","TXNAME"),
         keytype="GENEID"
      )
   );
   tx2geneMm10 <- renameColumn(tx2geneMm10,
      from=c("GENEID", "TXNAME", "TXTYPE"),
      to=c("gene_id", "transcript_id", "transcript_type"));

   # add gene_name using org.Mm.eg.db
   if (suppressPackageStartupMessages(require(org.Mm.eg.db))) {
      gene_ids <- values(genes(TxDb.Mmusculus.UCSC.mm10.knownGene))$gene_id;
      gene_namesL <- mget(gene_ids,
         org.Mm.egSYMBOL,
         ifnotfound=NA);
      ## Convert list to vector taking the first gene_name each
      ## (All genes should only have one SYMBOL but there is no
      ## hard constraint so we should make absolutely sure to
      ## use only one value per gene.)
      gene_names <- unlist(heads(S4Vectors::List(gene_namesL), 1));
      ## Replace NA with LOC# format
      ## Note we use gsub() to ensure the data fits the expected format
      if (any(is.na(gene_names))) {
         gene_na <- which(is.na(gene_names));
         gene_names[gene_na] <- gsub("^([0-9]+)$", "LOC\\1",
            names(gene_names[gene_na]));
      }
      tx2geneMm10$gene_name <- gene_names[as.character(tx2geneMm10$gene_id)];
   } else {
      ## If we have no gene annotations, use the gene_id values
      tx2geneMm10$gene_name <- as.character(tx2geneMm10$gene_id);
   }
   # print the first 20 rows to show the content
   print(head(tx2geneMm10, 20));
}
```

Next we flatten transcript exons to the gene level using
the function `splicejam::flattenExonsBy()`.

We also flatten exons to the transcript level, which has
the effect of combining CDS exons alongside non-CDS exons.
The result is used when plotting transcript exon structures.

In both cases, the argument `genes=c("Gria1", "Ntrk2")` is
used as a convenience to produce data for only these two genes.

> We recommend using detected transcripts here, with the
argument `detectedTx`.

```{r, flatten_exons}
if (require(TxDb.Mmusculus.UCSC.mm10.knownGene)) {
   # flatten exons to the gene level
   # for speed, we will only process "Gria1", and "Ntrk2"
   flatExonsByGeneMm10 <- flattenExonsBy(exonsByTx=exonsByTxMm10,
      cdsByTx=cdsByTxMm10,
      by="gene",
      genes=c("Gria1", "Ntrk2"),
      tx2geneDF=tx2geneMm10,
      verbose=FALSE);

   # to be fancy, also flatten transcripts, to include CDS ranges   
   flatExonsByTxMm10 <- flattenExonsBy(exonsByTx=exonsByTxMm10,
      cdsByTx=cdsByTxMm10,
      tx2geneDF=tx2geneMm10,
      by="tx",
      genes=c("Gria1", "Ntrk2"));
}
```

#### Plot gene-exon models using gene2gg()

Once the gene-transcript-exon data is prepared, the exon
structure can be plotted using the function `splicejam::gene2gg()`:

```{r, gene_plot, fig.height=5, fig.width=7}
if (require(TxDb.Mmusculus.UCSC.mm10.knownGene)) {
   # Pull out Gria1
   grlGria1 <- flatExonsByGeneMm10[["Gria1"]];

   ## Plot a basic gene-exon structure
   ggGria1exons <- gene2gg(gene="Gria1",
      flatExonsByGene=flatExonsByGeneMm10,
      exonLabelSize=6);
   print(ggGria1exons + ggtitle("Gria1 exons"));

   ## Compare to the gene structure without compressing introns
   gg1full <- gene2gg(gene="Gria1",
      flatExonsByGene=flatExonsByGeneMm10,
      compressGaps=FALSE)
   print(gg1full);
   
   ## Plot a slightly more detailed gene-transcript-exon structure
   ggGria1exonsTx <- gene2gg(gene="Gria1",
      flatExonsByGene=flatExonsByGeneMm10,
      flatExonsByTx=flatExonsByTxMm10,
      tx2geneDF=tx2geneMm10);
   print(ggGria1exonsTx + ggtitle("Gria1 (compressed introns)"));
   
   ## Notice how difficult it is to see exon15 and exon16 are
   ## mutually exclusive exons
   gg2full <- gene2gg(gene="Gria1",
      flatExonsByGene=flatExonsByGeneMm10,
      flatExonsByTx=flatExonsByTxMm10,
      tx2geneDF=tx2geneMm10,
      compressGaps=FALSE);
   print(gg2full + ggtitle("Gria1 (uncompressed introns)"));

}
```


## Coverage data

There are currently two main sources of coverage data:

* Files in `bigWig` format, accessible either as a file or web URL.
* GRanges objects, with one or more columns containing numeric coverage
data in the form of `NumericList`. This data is generated using
`rtracklayer::import(...,type="bigWig")` but it is recommended
to use the function `splicejam::getGRcoverageFromBw()`.

The next step is to define biological samples for each
set of coverage data. For this step, we use a `data.frame`
called `"filesDF"`, with these colnames:

* `"url"` - one of the following:
    * path to a local file
    * valid web URL
    * value from `colnames(values(covGR))` when `covGR` is supplied
    as a `GRanges` object containing coverage data.

* `"type"` - the type of file
    * `"bw"` for bigWig data
    * `"junction"` for a BED12 file containing splice junction reads
    * `"coverage_gr"` for coverage data provided in `covGR`

* `"sample_id"` - biological sample identifier, used to aggregate
files into the same plot panel.
* `"scale_factor"` - (optional) column with numeric values which are
multiplied by the score of the file. These values can help
normalize scores across samples.

### Coverage in bigWig format

A bigWig coverage file can only store coverage for one strand,
when using strand-specific RNA-seq. To provide negative strand data,
either provide bigWig coverage scores which are negative values
(never above zero, and containing at least one negative value),
or supply a negative value for the `"scale_factor"` field in
`filesDF`.

The example below defines a `data.frame` named `filesDF`
which contains bigWig coverage, and splice junction data.

```{r, files_df}
## assemble a data.frame
baseurl <- "https://orio.niehs.nih.gov/ucscview/farrisHub/mm10/";

# BED files with junction reads
bedext <- ".STAR_mm10.combinedJunctions.bed";
bwext <- c("492_1.sickle.merged.cutadapt.STAR_mm10.pos.bw",
   "492_1.sickle.merged.cutadapt.STAR_mm10.neg.bw");
c1 <- c("CA1", "CA2");
r1 <- c("CB", "DE");
bedsamples <- paste0(rep(c1, each=2), "_", r1);
bedurls <- paste0(baseurl,
   bedsamples,
   bedext);

# bigWig files with strand-specific read coverage
bwsamples <- paste0(rep(c1, each=4),
   rep(r1, each=2));
bwsamples1 <- paste0(rep(c1, each=4),
   "_",
   rep(r1, each=2));
bwurls <- paste0(baseurl, "NS50211",
   bwsamples,
   bwext);

# Assemble into a data.frame
filesDF <- data.frame(stringsAsFactors=FALSE,
   check.names=FALSE,
   url=c(bedurls, bwurls),
   type=rep(c("junction", "bw"), c(4,8)),
   sample_id=c(bedsamples, bwsamples1),
   scale_factor=rep(c(1,3), c(length(bedurls), length(bwurls))));
filesDF;

```

### Coverage in GRanges format

Example coverage data is provided as data, to demonstrate
the format expected. Note that in this case, there is no
coverage defined for introns. The bigWig workflow by default
adds gaps between exons, and retrieves coverage data for 
introns as well.

The `GRanges` object below uses an example set of exons,
`test_exon_wide_gr`, and corresponding coverage `GRanges`
is `test_exon_wide_gr`. Note that both share the same
names and genomic coordinates.

Ultimately, coverage data should be split by the same
`GRanges` used to define exons (and introns) for the
Sashimi plot.

```{r, test_cov_data}
data(test_exon_wide_gr);
test_exon_wide_gr;
data(test_cov_wide_gr);
test_cov_wide_gr;
```

### Plotting exon coverage data

Some basic plotting functions are provided, in fact they
are rolled into the Sashimi plot itself. They can help
visualize the coverage by itself.

The function `splicejam::exoncov2polygon()` converts coverage
to polygons, by default it returns a format convenient for
`geom_polygon()` or `ggforce::geom_shape()`, but if
`coord_style="base"` then it returns a matrix with blank
rows separating each polygon, suitable for R base graphics
using `graphics::polygon()`.

The argument `covNames` must contain one of more columns in
`colnames(values(test_cov_wide_gr))`.

```{r, cov_plots, fig.height=5, fig.width=7}
# To plot a simple GRanges object
widecovdf <- exoncov2polygon(test_cov_wide_gr, covNames="sample_A");

ggWide3 <- ggplot(widecovdf,
      aes(x=x, y=y, group=gr, fill=gr, color=gr)) +
   ggforce::geom_shape(alpha=0.7) +
   colorjam::theme_jam() +
   colorjam::scale_fill_jam() +
   colorjam::scale_color_jam();
print(ggWide3);
```

#### Compressing intron coordinates

The plot above demonstrates default genomic coordinates,
and shows how large intron regions are compared to typical
exons (in mammals anyway.) For better visualization of the
transcript exons, we compress introns using the
function `splicejam::make_ref2compresssed()`. This function
returns a list with several useful components, one of which
is `trans_grc` which is a ggplot2 transformation function.

The argument `nBreaks=10` defines 10 fixed x-axis label
positions, which by default use exon boundaries. To label
every exon boundary, use `nBreaks=Inf`.

The key step when using ggplot2 is to use the function
`ggplot2::scale_x_continuous()`, using argument
`trans=ref2c$trans_grc`. This argument defines the
x-axis compression, x-axis breaks, x-axis minor breaks,
and x-axis labels.

Alternatively, you can use `ggplot2::coord_trans(x=ref2c$trans_grc)`
however, that function only compresses the x-axis coordinates
and does not define breaks or labels.

```{r, compress_intron_cov, fig.height=5, fig.width=7}
# Now compress the introns keeping axis labels
ref2c <- make_ref2compressed(test_cov_wide_gr,
   nBreaks=10);

# Repeat ggplot using ref2c$trans_grc
ggWide3c <- ggWide3 +
   scale_x_continuous(trans=ref2c$trans_grc) +
   xlab("chr1 (compressed introns)") +
   ggtitle("exons (compressed introns)");
print(ggWide3c);
```

## Splice junction data

Splice junction data can be provided in two forms:

* Files in `BED6` or `BED12` format, accessible either as a file or web URL.
The `BED6` format uses the score column to count reads, and uses
the genomic coordinates to indicate the splice junction span. The
`BED12` format is inverted, so there is one base on each side, and
the junction region is in the middle. In this case, the one base is trimmed
off each side, and the internal region is used to indicate the junction
spanning region.
* GRanges objects, where the `"score"` column indicates the number of
splice junction reads, and the `"sample_id"` column indicates the
biological sample. Therefore all splice junctions for all samples
should be condensed into one `GRanges` object.

Currently, the `rtracklayer::import()` function does not support
`bigBed` format, and there is no apparent plan to provide that
capability. It is recommended to use the UCSC tool `bigBedToBed`
to convert `bigBed` files to BED format.

A set of sample splice junction data is provided to show the
expected format of junction data. Note it has two columns
`"score"` and `"sample_id"`.

```{r, test_junc}
data(test_junc_wide_gr);
test_junc_wide_gr;
```

### Plotting splice junction data

Some basic plotting functions are provided, which are
rolled into the Sashimi plot. They can help
visualize splice junction data directly, as needed.

The key function here is `splicejam::grl2df(...,shape="junction")`
which configures wide junction arcs suitable for
`splicejam::geom_diagonal_wide_arc()`. (Note that
`ggplot2::geom_polygon()` would also work.)

```{r, plot_test_junc, fig.height=5, fig.width=7}
# To plot junctions, use grl2df(..., shape="junction")
junc_wide_df <-grl2df(test_junc_wide_gr,
   shape="junction");

ggWide1 <- ggplot(junc_wide_df,
      aes(x=x, y=y, group=gr_name, fill=gr_name, color=gr_name)) +
   splicejam::geom_diagonal_wide_arc() +
   colorjam::theme_jam() +
   colorjam::scale_fill_jam(alpha=0.7) +
   colorjam::scale_color_jam() +
   xlab("chr1") +
   ggtitle("junctions (full intron width)")
print(ggWide1);
```

For more insight into creating a Sashimi plot manually,
see the help for data `splicejam::test_junc_wide_gr`.

## Prepare Sashimi Data

The main function `prepareSashimi()` is intended to assemble the
sources of data together into one list of `data.frame` results
which can be used to produce ggplot2 visualizations.

### Data obtained from external files

For this example, we use the Gria1 gene in mouse mm10,
pulling from a web-accessible track hub. Note that
the `bigBed` files were already converted to `BED12`
format using the UCSC tool `bigBedToBed`.


```{r, prepare_sashimi}
shGria1 <- prepareSashimi(gene="Gria1",
   flatExonsByGene=flatExonsByGeneMm10,
   minJunctionScore=100,
   sample_id=c("CA1_CB", "CA2_CB"),
   filesDF=filesDF)
```

### Data provided using GRanges objects

Alternatively, you can provide data in the form of
`GRanges` objects, and not point to specific files.

Note that `filesDF` is still required for coverage
in `GRanges` format, in order to define the
`"sample_id"` for each coverage.

```{r, prepare_sashimi_gr}
data(test_exon_wide_gr);
data(test_junc_wide_gr);
data(test_cov_wide_gr);
testfilesDF <- data.frame(url="sample_A",
   type="coverage_gr",
   sample_id="sample_A",
   scale_factor=1);
testfilesDF;

# Now prepare sashimi data
sh1 <- prepareSashimi(GRangesList(TestGene1=test_exon_gr),
   filesDF=testfilesDF,
   gene="TestGene1",
   sample_id="sample_A",
   covGR=test_cov_gr,
   juncGR=test_junc_gr);

```

At this point, we can just plot the result.

```{r, plot_sashimi_gr, fig.height=5, fig.width=7}
# Plot sample Sashimi data
plotSashimi(sh1);
```

## Plot Sashimi Data

Once the Sashimi data is prepared, the `splicejam::plotSashimi()`
function is used to create a visualization that
can be customized in several ways.

The output of `splicejam::plotSashimi()` is a `ggplot`
object, which can be printed to create a plot, or
can be modified as needed.

```{r, plot_sashimi, fig.height=10, fig.width=8}
ggGria1 <- plotSashimi(shGria1,
   junc_color=alpha2col("goldenrod1", 0.4),
   junc_fill=alpha2col("goldenrod1", 0.4),
   show=c("coverage", "junction", "junctionLabels"),
   fill_scheme="sample_id");
print(ggGria1);
```


### Adding gene-exon model to a Sashimi plot

To add a gene-exon model, and even transcript-exon
models, we will use the `cowplot` package to arrange
multiple `ggplot` objects together.

Some guidelines:

* This step requires the `cowplot` package.
* Sashimi plots should have facets with one column,
for example ` + facet_wrap(~sample_id, ncol=1)`.
* The Sashimi plot and the gene-exon plot should
share identical x-axis range and axis breaks.
* The x-axis labels of Sashimi plot panels should
be hidden, since the gene-exon plot will display
the genomic coordinates.

The example below demonstrates combining a gene-exon
plot with a Sashimi plot, meeting the required
guidelines.

```{r, plot_sashimi_exons, fig.height=8, fig.width=7}
if (suppressPackageStartupMessages(require(cowplot))) {
   cpGria1 <- cowplot::plot_grid(
      ggGria1 + theme(axis.text.x=element_blank()) + xlab(NULL),
      ggGria1exonsTx + ggtitle(NULL) + expand_limits(y=-2),
      ncol=1,
      align="v", 
      axis="lr",
      rel_heights=c(4,4));
   print(cpGria1);
}
```

### Sashimi plot with unscaled introns

```{r plot_sashimi_wide, fig.height=10, fig.width=7}
if (require(cowplot)) {
   cpWide <- cowplot::plot_grid(ncol=1,
      ggGria1 + scale_x_continuous() + theme(axis.text.x=element_blank()) + xlab(NULL),
      gg2full,
      axis="lr",
      align="v");
   print(cpWide);
}
```

### Zoom into specific exons

There are two ways to zoom into a region of interest:

1. Prepare one Sashimi panel, and use `ggforce::facet_zoom()`
2. Prepare Sashimi data for a specific set of exons.

#### Zoom one panel using facet_zoom

The example below shows the `ggforce::facet_zoom()`
capability.

Note the most important step is to transform the x-axis
values when using `xlim` to subset the range.
The `splicejam::prepareSashimi()` function returns
`x$ref2c$transform()` which transforms genomic
coordinates into compressed coordinates.

```{r, zoom_sashimi, fig.height=5, fig.width=7}
plotSashimi(sh1) + 
   facet_zoom(xlim=sh1$ref2c$transform(c(370, 410)));
```

#### Zoom multiple panels

In order to zoom multiple panels, a few steps must be
followed:

* the data for each panel must be prepared separately,
* the `ggforce::facet_zoom()` method above is applied to
each plot
* the plots are arranged together using `cowplot::plot_grid()`.

```{r, zoom_gria1, fig.height=10, fig.width=12}
zoomStart <- start(flatExonsByGeneMm10[["Gria1"]]["Gria1_exon14"]);
zoomEnd <- end(flatExonsByGeneMm10[["Gria1"]]["Gria1_exon18a"]);

# Now prepare sashimi data
shGria1L <- lapply(c("CA1_CB", "CA2_CB"), function(iSample){
   # devtools::load_all("/Users/wardjm/Projects/Ward/splicejam")
   shX <- prepareSashimi(gene="Gria1",
      flatExonsByGene=flatExonsByGeneMm10,
      minJunctionScore=100,
      sample_id=iSample,
      doStackJunctions=FALSE,
      filesDF=filesDF);
});
ggGriaL <- lapply(shGria1L, function(shX){
      ggX <- plotSashimi(shX) +
         facet_zoom(xlim=shX$ref2c$transform(c(zoomStart, zoomEnd)));
});

# now use cowplot
if (require(cowplot)) {
   cpMulti <- do.call(cowplot::plot_grid, ggGriaL);
   print(cpMulti);
}


```

